---
title: "Task 4 : Examining R packages"
output: html_notebook
---

The following lists the R packages available in this machine with one line descriptions.
```{r}
library()
```

The following lists the R packages currently attached/loaded to this session.
```{r}
search()
```
The following opens up the help page for how to use the `vignette` function.
```{r}
?vignette
```

Installing the `seqinr` package.
```{r}
install.packages("seqinr",repos="https://cran.rstudio.com/")

# code below is a good way to avoid producing errors when a package has not been installed 
# if (! requireNamespace("seqinr", quietly=TRUE)) {
#   install.packages("seqinr")
# }
```

Loading and examining the package.
```{r}
library(seqinr)
library(help="seqinr") # pulls up information about the package
ls("package:seqinr") # lists all functions available in this package
```

The command below pulls up the help page for the function `a`.
It seems like it converts three letter amino-acid codes into their single letter expression.
```{r}
?seqinr::a
```

Testing out the function for Tyrosine. Looks about right!
```{r}
seqinr::a("Tyr")
```

Testing the `words` function.
It returned all possible combinations permutations of the items in the input vector of length 3.
```{r}
seqinr::words(3, c("A", "G", "C", "U"))
```

The command below pulled up the help page for the `data` function provided in `utils`.
It seems like it loads specified datasets OR lists any that are available.
```{r}
?data
```

Checking what data sets are available in the `seqinr` package.
```{r}
data(package="seqinr")
```

Loading one of the data sets provided in `seqinr` called `aaindex`.
```{r}
data(aaindex, package="seqinr")   
```

Pulling up the help documentation for the `aaindex` data set.
Seems like a list of the physicochemical & biological properties of the 20 AAs.
It is a list of 544 elements, indexed by aaindex accession number.
Seems very organized and thorough!
```{r}
?aaindex
```

Examining some indices...
```{r}
aaindex$FASG890101
aaindex$PONJ960101
```
Plotting the data. 
```{r}
plot(aaindex$FASG890101$I, # setting x values to be hydrophobicity
     aaindex$PONJ960101$I, # setting y values to be volume
     xlab="hydrophobicity", ylab="volume", type="n")
text(aaindex$FASG890101$I,
     aaindex$PONJ960101$I,
     labels=a(names(aaindex$FASG890101$I))) # letting labels be the single letter AA expression using the function a
```

Exploring some plotting code provided.
Read through each function's documentation to understand what is going on. (annotations in comments)
```{r}
seqinr::choosebank("swissprot") # creates connection to remote database
mySeq <- seqinr::query("mySeq", "N=MBP1_YEAST") # getting list of seq names from ACNUC : args are 1) name of list and 2) seqname (ID/locus using N tag)
mbp1 <- seqinr::getSequence(mySeq) # getting sequence from the qaw object returned by function query, returns a vector of single, upper case characters
seqinr::closebank() # closing connection
x <- seqinr::AAstat(mbp1[[1]]) # getting protein seq information using a protein seq, composistion (count) and properties, ALSO plots presence of residues based on character bc plot argument is set to TRUE by default (colorful plot)
barplot(sort(x$Compo), cex.names = 0.6) # plotting the count of each residue
```

The `sos` package is helpful in finding R packages.
```{r}
if (! requireNamespace("sos", quietly=TRUE)) {
    install.packages("sos")
}
library(help = sos)       # basic information
browseVignettes("sos")    # available vignettes
 
sos::findFn("moving average") # looks for functions related to query "moving average"
```
