---
title: "BCB420 A1"
author: "Izumi Ando"
output: 
  html_document:
    toc: true
    toc_depth: 2
bibligraphy: A1_references.bib

---

# 0 - Dataset Overview
The dataset I selected is called "SF3B1 promotes glycolysis and tumor malignancy through splicing-independent activation of HIF1α [RNA-Seq]" (GEO accession: GSE201427).

# 1 - Download the Data

```{r eval=FALSE}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("GEOquery")
```

```{r}
# getting the GEO description
geoID <- "GSE201427"
# figure out a way to access the cache data if available 
gse <- GEOquery::getGEO(geoID, GSEMatrix=FALSE)
```

```{r eval=FALSE}
library("GEOquery")

# getting the GEO description
geoID <- "GSE201427"
# figure out a way to access the cache data if available 
gse <- GEOquery::getGEO(geoID, GSEMatrix=FALSE)
# gse@header$summary
suppFiles <- GEOquery::getGEOSuppFiles(geoID, fetch_files=FALSE)
# we can see that there is only one file available (from the env tab)
dataFile <- suppFiles$fname[1]

# only downloading dataset if it is not available
dir <- file.path(getwd())
dataFilePath <- file.path(dir, geoID, dataFile)
if(!file.exists(dataFilePath)){
  dataFileDownload <- GEOquery::getGEOSuppFiles(geoID, 
                                                filter_regex = dataFile, 
                                                baseDir = dir, 
                                                fetch_files = TRUE)
}

# reading in the data
panc1Data <- read.table(dataFilePath, header = TRUE, check.names = TRUE)
dim(panc1Data)
# getting issues about nul characters. it looks fine on 
```

```{r}
# getting issues about nul characters. it looks fine on excel UI.
# using readxl package https://readxl.tidyverse.org/reference/read_excel.html

# checking if tidyverse is installed
"tidyverse" %in% rownames(installed.packages())
# install.packages("tidyverse") 
library("tidyverse")
# this package is included in tidyverse but needs to be loaded explicitly bc it is not core
library("readxl") 

panc1Data2 <- readxl::read_xlsx(dataFilePath) 
# this worked! it seemed like xlsx has special encoding

# not understanding the differences between gse, gpl. and gsm
# https://www.bioconductor.org/packages/release/bioc/vignettes/GEOquery/inst/doc/GEOquery.html
```

### Information About the Series (GSE)
Title: `r gse@header$title`  
GEO Accession: `r gse@header$geo_accession`   
Submission Date: `r gse@header$submission_date`   
Last Update: `r gse@header$last_update_date`   
Associated Publication PubMed ID: `r gse@header$pubmed_id`
Contact Information:  
&nbsp;&nbsp; Contact Name: `r gse@header$contact_name`  
&nbsp;&nbsp; Contact Institute: `r gse@header$contact_institute`  
&nbsp;&nbsp; Contact Name: `r gse@header$contact_email`  

### Information About the Platform (GPL)

```{r eval=FALSE}
# seeing the distribution of the data
# code taken from lecture5_normalization
counts_density <- apply(log2(panc1Data2[3:8]), 2, density)

xlim <- 0
ylim <- 0
for (i in 1:length(counts_density)){
  xlim <- range(c(xlim, counts_density[[i]]$x))
  ylim <- range(c(ylim, counts_density[[i]]$y))
}

cols <- rainbow(length(counts_density))
ltys <- rep(1, length(counts_density))

plot(counts_density[[1]], xlim=xlim, ylim=ylim, type="n", 
     ylab="Smoothing density of log2-CPM",
     main="",
     cex.lab=0.85)
for (i in 1:length(counts_density)){
  lines(counts_density[[i]], col=cols[i], ltys=ltys[i])
}
```

## 2 - Assessing the Data
```{r}
# THIS WORKS AND IS FORMATTED
# MAKE SURE TO CITE THE SLIDES

# removing extra columns for different O2 exposure levels
panc1Data2_sub <- panc1Data2[, c(1, 2, 5, 6, 7, 8, 9, 10)]

# Extract only numeric data (third column onwards)
numeric_data <- panc1Data2_sub[, 3:ncol(panc1Data2_sub)]


# Calculate density for each sample (column-wise)
#counts_density <- apply(log2(numeric_data), 2, density)
# realized this data is already log cpm
counts_density <- apply(numeric_data, 2, density)

# Determine x and y limits across all samples
xlim <- range(sapply(counts_density, function(d) d$x))
ylim <- range(sapply(counts_density, function(d) d$y))

# Assign colors to each sample
cols <- rainbow(length(counts_density))
ltys <- rep(1, length(counts_density))

# Initialize an empty plot
plot(counts_density[[1]], xlim=xlim, ylim=ylim, type="n",
     xlab="log-CPM", ylab="Density",
     main="Density Plot of log-CPM in PANC-1 RNA-seq",
     cex.lab=0.85)

# Add density curves for each sample
for (i in 1:length(counts_density)) {
  lines(counts_density[[i]], col=cols[i], lty=ltys[i])
}

# Add legend (using column names from numeric data)
legend("topright",
       legend=colnames(numeric_data),
       col=cols, lty=ltys, cex=0.75, ncol=1,
       border="blue", text.col="green4",
       merge=TRUE, bg="gray90")
```

```{r}
# taken from lecture5_normalization
# using data from box above
boxplot(numeric_data, xlab = "Samples", ylab = "log-CPM",
        las = 2, cex = 0.75, cex.lab = 0.75,
        cex.axis = 0.75, main = "Box Plot of PANC-1 RNA-seq")

# median for each box
abline(h = median(apply(numeric_data, 2, median)),
       col = "purple", lwd = 1.0, lty = "dashed")
```

```{r}
panc1Data <- panc1Data2
```

### Statistics About the Dataset
**Number of rows:** `r nrow(panc1Data)`  
**Number of Entrez IDs:** `r nrow(panc1Data) - nrow(panc1Data[panc1Data$EntrezID == "NA", ])`   
**Number of Gene Names:** `r nrow(panc1Data) - nrow(panc1Data[panc1Data$Gene == "NA", ])`    

### Other points to note:   
* The sample labels (column names) are formatted the the following way "A_<NUM>_Panc1_<EXP>_<O2>" where <NUM> is the sample number, <EXP> is the experiment group either control siRNA (siCtrl) or SF3B1 targeting siRNA (siSF3B1), and <O2> is the oxygen level (Hx for hypoxia or 1%, and Nx for 21%). The original data file seems to only include 8 samples although the GSE201427 description states that there are 9. However, the missing sample seems to be **A_7_Panc1_siCtrl_Nx** which is not part of the experiment I will be focusing on so it is not an issue.
* The values in the dataset is already in log-CPM

```{r}
# modified code from lecture 4 get the data 
list_of_samples <- gse@gsms
samples_type <- do.call(rbind, 
                        lapply(list_of_samples,
                               FUN = function(x){c(x@header$title, 
                                                   x@header$characteristics_ch1)
                                 }))
colnames(samples_type) <- c("Sample Name", "Oxygen Exposure", 
                            "siRNA Treatment", "Cell Line")

samples_type[, "Oxygen Exposure"] <- gsub(samples_type[, "Oxygen Exposure"],
                                          pattern = "condition: ",
                                          replacement = "")

samples_type[, "siRNA Treatment"] <- gsub(samples_type[, "siRNA Treatment"],
                                          pattern = "treatment: ",
                                          replacement = "")

samples_type[, "Cell Line"] <- gsub(samples_type[, "Cell Line"],
                                          pattern = "cell line: ",
                                          replacement = "")

samples_type <- as.data.frame(samples_type)
```

## 3 - Identifier mapping
```{r eval=FALSE}
# taking code from lecture 5 identifier mapping
# BiocManager::install("biomaRt")
library("biomaRt")
biomaRt::listMarts()
ensembl <- biomaRt::useMart("ensembl")
# mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
biomaRt::listDatasets(ensembl) # need to use hsapiens_gene_ensembl
mart <- biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl")
panc1_ids <- panc1Data[, 1]

# conversion <- biomaRt::getBM(
#     attributes = c("entrezgene_id", "hgnc_symbol"),
#     filters = "entrezgene_id",
#     values = panc1_ids,
#     mart = mart
# )
# the code above errors, i think biomart only has 3794 entrez ids
searchAttributes(mart=mart, "entrez")
```
```{r}
# found new tool org.Hs.eg.db (annotation package)
# BiocManager::install("org.Hs.eg.db")
# https://bioconductor.org/packages/release/data/annotation/html/
# docs: https://bioconductor.org/packages/release/data/annotation/manuals/org.Hs.eg.db/man/org.Hs.eg.db.pdf
library("org.Hs.eg.db")
# AnnotationDbi is a dependency of org.Hs.eg.db
AnnotationDbi::keytypes(org.Hs.eg.db)
AnnotationDbi::columns(org.Hs.eg.db)
v_panc1_ids <- as.vector(panc1_ids)
v_panc1_ids <- as.character(unlist(v_panc1_ids))
mapped <- AnnotationDbi::select(org.Hs.eg.db, keys=v_panc1_ids, columns=c("ENTREZID", "SYMBOL"), keytype = "ENTREZID")

nrow(mapped[mapped$SYMBOL == "NA", ]) # number of NA's from the mapping
nrow(panc1Data[panc1Data$Gene == "NA", ]) # number of NA's from the original data 

# check to see if there are things that can be supplemented
joined <- cbind(mapped, panc1Data[, 2])
nrow(joined[joined$Gene == "NA" & joined$SYMBOL == "NA", ])
# this tells us that the original data set has more gene symbol names 
# DECISION: NOT MAPPING
```
```{r}
# removing all rows without hugo ids
mapped_panc1Data <- panc1Data[panc1Data$Gene != "NA", ]
geneCol <- mapped_panc1Data$Gene
mapped_panc1Data <- mapped_panc1Data[, 3:length(mapped_panc1Data)]
rownames(mapped_panc1Data) <- geneCol
```


## 4 - Cleaning the Data

* min number of samples is 3

```{r}
# code taken from lecture 5 normalizing our dataset
min_num_samples <- 3
# removing log from data values 
cpm_non_log <- 2^mapped_panc1Data
data_matrix <- as.matrix(cpm_non_log)
keep <- rowSums(data_matrix > 1) > min_num_samples
filtered_data_matrix <- data_matrix[keep,]

filtered_cpm_log2 <- log(filtered_data_matrix, base = 2)

# Calculate density for each sample (column-wise)
#counts_density <- apply(log2(numeric_data), 2, density)
# realized this data is already log cpm
counts_density <- apply(filtered_cpm_log2, 2, density)

# Determine x and y limits across all samples
xlim <- range(sapply(counts_density, function(d) d$x))
ylim <- range(sapply(counts_density, function(d) d$y))

# Assign colors to each sample
cols <- rainbow(length(counts_density))
ltys <- rep(1, length(counts_density))

# Initialize an empty plot
plot(counts_density[[1]], xlim=xlim, ylim=ylim, type="n",
     xlab="log-CPM", ylab="Density",
     main="Density Plot of log-CPM in filtered PANC-1 RNA-seq",
     cex.lab=0.85)

# Add density curves for each sample
for (i in 1:length(counts_density)) {
  lines(counts_density[[i]], col=cols[i], lty=ltys[i])
}

# Add legend (using column names from numeric data)
legend("topright",
       legend=colnames(numeric_data),
       col=cols, lty=ltys, cex=0.75, ncol=1,
       border="blue", text.col="black",
       merge=TRUE, bg="gray90")

```

## 4 - Normalization

*realizing now that the next portion only goes to CPM, which we already have so we might be able to omit

to do
* do the MDS plot
* see if there is anything beyond that
* see if i should use a diff normalization or if i can skip all together

```{r}
library("edgeR")

# code taken from lecture 5 normalizing our dataset
d <- edgeR::DGEList(counts=filtered_data_matrix, group=samples_type$"siRNA Treatment")
d <- edgeR::calcNormFactors(d)
normalized_counts <- cpm(d)

log_new <- log(normalized_counts, base = 2)

# Calculate density for each sample (column-wise)
#counts_density <- apply(log2(numeric_data), 2, density)
# realized this data is already log cpm
counts_density <- apply(log_new, 2, density)

# Determine x and y limits across all samples
xlim <- range(sapply(counts_density, function(d) d$x))
ylim <- range(sapply(counts_density, function(d) d$y))

# Assign colors to each sample
cols <- rainbow(length(counts_density))
ltys <- rep(1, length(counts_density))

# Initialize an empty plot
plot(counts_density[[1]], xlim=xlim, ylim=ylim, type="n",
     xlab="log-CPM", ylab="Density",
     main="Density Plot of log-CPM in normalized PANC-1 RNA-seq",
     cex.lab=0.85)

# Add density curves for each sample
for (i in 1:length(counts_density)) {
  lines(counts_density[[i]], col=cols[i], lty=ltys[i])
}

# Add legend (using column names from numeric data)
legend("topright",
       legend=colnames(numeric_data),
       col=cols, lty=ltys, cex=0.75, ncol=1,
       border="blue", text.col="black",
       merge=TRUE, bg="gray90")
```